Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2023-03-01 paper/main.tex
edits up through end of results

% supplemental figures
\newcommand{\topicWeights}{1}
\newcommand{\individualKnowledgeMapsA}{2}
\newcommand{\individualKnowledgeMapsB}{3}
\newcommand{\individualKnowledgeMapsC}{4}
\newcommand{\individualLearningMapsA}{5}
\newcommand{\individualLearningMapsB}{6}

@@

    \caption{\textbf{Modeling course content.} \textbf{A. Building a document pool
    from sliding windows of text.} We decompose each lecture's transcript into a
    series of overlapping sliding windows. The full set of transcript snippets (across
    all windows) may be treated as a set of ``documents'' for training a text
    embedding model. \textbf{B. Constructing lecture content
    \textit{trajectories}.} After training our model on the sliding windows from
    both lectures, we transform each lecture into a ``trajectory'' through text
    embedding space by joining the embedding coordinates of successive sliding
    windows parsed from its transcript. \textbf{C. Embedding multiple lectures and
    questions in a shared space.} We apply the same model (trained on the two
    lectures' windows) to both lectures, along with the text of each question in
    our pool (Supplementary Tab.~\questions), to project them into a shared text embedding space.
    This results in one trajectory per lecture and one coordinate for each
    question. Here, we have projected the 15-dimensional embeddings onto their first
    3 principal components for visualization.}

@@

To study in detail how participants' conceptual knowledge changed over the
course of the experiment, we first sought to model the conceptual content
presented to them at each moment throughout each of the two lectures. We
adapted an approach we developed in prior work~\citep{HeusEtal21} to identify
the latent themes in the lectures using a topic model~\citep{BleiEtal03}.
Briefly, topic models take as input a collection of text documents, and learn a
set of ``topics'' (i.e., latent themes) from their contents. Once fit, a topic
model can be used to transform arbitrary (potentially new) documents into sets
of ``topic proportions,'' describing the weighted blend of learned topics
reflected in their texts. We parsed automatically generated transcripts of the
two lectures into overlapping sliding windows, where each window contained the
text of the lecture transcript from a particular time span. We treated the set
of text snippets (across all of these windows) as documents to fit our model
(Fig.~\ref{fig:sliding-windows}A; see \nameref{subsec:topic-modeling}).
Transforming the text from every sliding window with our model yielded a
number-of-windows by number-of-topics (15) topic-proportions matrix
describing the unique mixture of broad themes from both lectures reflected in
each window's text. Each window's ``topic vector'' (i.e., column of the
topic-proportions matrix) is analogous to a coordinate in a 15-dimensional space whose axes
are topics discovered by the model. Within this space, each lecture's sequence
of topic vectors (i.e., corresponding to its transcript's overlapping text
snippets across sliding windows) forms a \textit{trajectory} that captures how
its conceptual content unfolds over time (Fig.~\ref{fig:sliding-windows}B). We
resampled these trajectories to a resolution of one topic vector for each
second of video (i.e., 1~Hz).

@@

Additionally, if we are able to accurately estimate a participant's knowledge about the
content tested by a given question, our estimates of their knowledge should carry some
predictive information about whether the participant is likely to answer that
question correctly or incorrectly. To further validate our knowledge estimates, we developed a statistical approach to test this claim.
For each question in turn, we used equation~\ref{eqn:prop} to estimate each
participant's knowledge at the given question's embedding space coordinate,
using all \textit{other} questions that participant answered on the same quiz.
%For each question in turn, for each
%participant, we used Equation~\ref{eqn:prop} to estimate (using all
%\textit{other} questions from the same quiz, from the same participant) the
%participant's knowledge at the held-out question's embedding coordinate.
For
each quiz, we grouped these estimates into two distributions: one for the
estimated knowledge at the coordinates of \textit{correctly} answered
questions, and another for the estimated knowledge at the coordinates of
\textit{incorrectly} answered questions (Fig.~\ref{fig:predictions}). We then
used independent samples $t$-tests to compare the means of these distributions
of estimated knowledge.

@@

Knowledge estimates need not be limited to the content of the lectures. As
illustrated in Figure~\ref{fig:knowledge-maps}, our general approach to
estimating knowledge from a small number of quiz questions may be applied to
\textit{any} content, given its text embedding coordinate. To visualize how
knowledge ``spreads'' through text embedding space to content beyond the
lectures participants watched, we first fit a new topic model to the lectures'
sliding windows with (up to)~$k = 100$~topics.
Conceptually, increasing the number of topics used by the model functions to
increase the ``resolution'' of the embedding space, providing a greater ability to
estimate knowledge for content that is highly similar to (but not precisely the same as)
that contained in the lectures.
Aside from increasing the number of topics from 15 to 100, all other
procedures and model parameters were carried over from the preceding analyses.
As in our other analyses, we resampled each lecture's topic trajectory to 1~Hz
and projected each question into a shared text embedding space.

@@

    \caption{\textbf{Mapping out the geometry of knowledge and learning.}
    \textbf{A. Average ``knowledge maps'' estimated using each quiz.} Each map
    displays a 2D projection of the estimated knowledge about the content
    reflected by \textit{all} regions of topic space (see
    \nameref{subsec:knowledge-maps}). The topic trajectories of each
    lecture are indicated by dotted lines, and the coordinates of each question are indicated by
    dots. Each map reflects an average across all participants. For
    individual participants' maps, see
    Supplementary Figures~\individualKnowledgeMapsA,~\individualKnowledgeMapsB,
    and~\individualKnowledgeMapsC. \textbf{B. Average ``learning maps''
    estimated between each successive pair of quizzes.} The learning maps follow
    the same general format as the knowledge maps in Panel A, but here the shading at each
    coordinate indicates the \textit{difference} between
    the corresponding coordinates in the indicated \textit{pair} of knowledge
    maps---i.e., how much the estimated knowledge ``changed'' between the two
    quizzes. Each map reflects an average across all participants. For
    individual participants' maps, see Supplementary
    Figures~\individualLearningMapsA~and~\individualLearningMapsB. \textbf{C.
    Word clouds for sampled points in topic space.} Each word cloud displays
    the relative weights of each word (via their relative sizes) reflected by the blend of topics
    represented at the locations of the stars on the maps. The words' colors
    indicate how much each word is weighted, on average, across all timepoints'
    topic vectors in the \textit{Four Fundamental Forces} (blue) and
    \textit{Birth of Stars} (green) videos, respectively.}

@@

We projected the resulting 100-dimensional topic vectors (for each second of
video and for each question) onto a shared 2-dimensional plane (see
\nameref{subsec:knowledge-maps}). Next, we sampled points from a $100 \times
100$ grid of coordinates that evenly tiled a rectangle enclosing the 2D
projections of the videos and questions. We used
Equation~\ref{eqn:rbf-knowledge} to estimate participants' knowledge at each of
these 10,000 sampled locations, and averaged these estimates across
participants to obtain an estimated average \textit{knowledge map}
(Fig.~\ref{fig:knowledge-maps}A). Intuitively, the knowledge map constructed
from a given quiz's responses provides a visualization of how ``much''
participants know about any content expressible by the fitted text embedding
model.

Several features of the resulting knowledge maps are worth noting. The average
knowledge map estimated from Quiz~1 responses (Fig.~\ref{fig:knowledge-maps}A,
leftmost map) shows that participants tended to have relatively little
knowledge about any parts of the text embedding space (i.e., the shading is
relatively dark everywhere). The knowledge map estimated from Quiz~2 responses
shows a marked increase in knowledge on the left side of the map (around
roughly the same range of coordinates traversed by the \textit{Four Fundamental
Forces} lecture, indicated by the dotted blue line). In other words,
participants' estimated increase in knowledge is localized to conceptual
content that is nearby (i.e., related to) the content from the lecture they
watched prior to taking Quiz~2. This localization is non-trivial: the knowledge
estimates are informed only by the embedded coordinates of the \textit{quiz
questions}, not by the embeddings of either lecture
(see Eqn.~\ref{eqn:rbf-knowledge}). Finally, the knowledge map estimated from Quiz
3 responses shows a second increase in knowledge, localized to the region
surrounding the embedding of the \textit{Birth of Stars} lecture participants
watched immediately prior to taking Quiz~3.

Another way of visualizing these content-specific increases in knowledge
after participants viewed each lecture is displayed in
Figure~\ref{fig:knowledge-maps}B. Taking the point-by-point difference between
the knowledge maps estimated from responses to a successive pair of quizzes
yields a \textit{learning map} that describes the \textit{change} in knowledge
estimates from one quiz to the next. These learning maps highlight that the
estimated knowledge increases we observed across maps were specific to the
regions around the embeddings of each lecture in turn.

Because the 2D projection we used to construct the knowledge and learning maps
is invertible, we may gain additional insights into these maps' meaning
by reconstructing the original high-dimensional topic vector for any location
on the map we are interested in. For example, this could serve as a useful
tool for an instructor looking to better understand which content areas a
student (or a group of students) knows well (or poorly). As a demonstration, we
show the top-weighted words from the blends of topics reconstructed from three
example locations on the maps (Fig.~\ref{fig:knowledge-maps}C): one point near
the \textit{Four Fundamental Forces} embedding (yellow); a second point near
the \textit{Birth of Stars} embedding (orange), and a third point
between the two lectures' embeddings (pink). As shown in the word clouds in the
Panel, the top-weighted words at the example coordinate near the \textit{Four
Fundamental Forces} embedding also tended to be weighted heavily by the topics
expressed in that lecture. Similarly, the top-weighted words at the example
coordinate near the \textit{Birth of Stars} embedding tended to be weighted
most heavily by the topics expressed in \textit{that} lecture. And the
top-weighted words at the example coordinate between the two lectures'
embeddings show a roughly even mix of words most strongly associated with each
lecture.

>>>>

% supplemental figures
\newcommand{\topicWordWeights}{1}
\newcommand{\topicWeights}{2}
\newcommand{\individualKnowledgeMapsA}{3}
\newcommand{\individualKnowledgeMapsB}{4}
\newcommand{\individualKnowledgeMapsC}{5}
\newcommand{\individualLearningMapsA}{6}
\newcommand{\individualLearningMapsB}{7}

@@

    \caption{\textbf{Modeling course content.} \textbf{A. Building a document pool
    from sliding windows of text.} We decompose each lecture's transcript into a
    series of overlapping sliding windows. The full set of transcript snippets (across
    all windows) may be treated as a set of ``documents'' for training a text
    embedding model. \textbf{B. Constructing lecture content
    \textit{trajectories}.} After training the model on the sliding windows from
    both lectures, we transform each lecture into a ``trajectory'' through text
    embedding space by joining the embedding coordinates of successive sliding
    windows parsed from its transcript. \textbf{C. Embedding multiple lectures and
    questions in a shared space.} We apply the same model (trained on the two
    lectures' windows) to both lectures, along with the text of each question in
    our pool (Supplementary Tab.~\questions), to project them into a shared text embedding space.
    This results in one trajectory per lecture and one coordinate for each
    question. Here, we have projected the 15-dimensional embeddings onto their first
    3 principal components for visualization.}

@@

To study in detail how participants' conceptual knowledge changed over the
course of the experiment, we first sought to model the conceptual content
presented to them at each moment throughout each of the two lectures. We
adapted an approach we developed in prior work~\citep{HeusEtal21} to identify
the latent themes in the lectures using a topic model~\citep{BleiEtal03}.
Briefly, topic models take as input a collection of text documents, and learn a
set of ``topics'' (i.e., latent themes) from their contents. Once fit, a topic
model can be used to transform arbitrary (potentially new) documents into sets
of ``topic proportions,'' describing the weighted blend of learned topics
reflected in their texts. We parsed automatically generated transcripts of the
two lectures into overlapping sliding windows, where each window contained the
text of the lecture transcript from a particular time span. We treated the set
of text snippets (across all of these windows) as documents to fit the model
(Fig.~\ref{fig:sliding-windows}A; see~\nameref{subsec:topic-modeling}).
Transforming the text from every sliding window with the model yielded a
number-of-windows by number-of-topics (15) topic-proportions matrix
describing the unique mixture of broad themes from both lectures reflected in
each window's text. Each window's ``topic vector'' (i.e., column of the
topic-proportions matrix) is analogous to a coordinate in a 15-dimensional space whose axes
are topics discovered by the model. Within this space, each lecture's sequence
of topic vectors (i.e., corresponding to its transcript's overlapping text
snippets across sliding windows) forms a \textit{trajectory} that captures how
its conceptual content unfolds over time (Fig.~\ref{fig:sliding-windows}B). We
resampled these trajectories to a resolution of one topic vector for each
second of video (i.e., 1~Hz).

@@

If we are able to accurately estimate a participant's knowledge about the
content tested by a given question, our estimates of their knowledge should carry some
predictive information about whether the participant is likely to answer that
question correctly or incorrectly. We developed a statistical approach to test this claim.
For each question in turn, we used equation~\ref{eqn:prop} to estimate each
participant's knowledge at the given question's embedding space coordinate,
using all \textit{other} questions that participant answered on the same quiz.
%For each question in turn, for each
%participant, we used Equation~\ref{eqn:prop} to estimate (using all
%\textit{other} questions from the same quiz, from the same participant) the
%participant's knowledge at the held-out question's embedding coordinate.
For
each quiz, we grouped these estimates into two distributions: one for the
estimated knowledge at the coordinates of \textit{correctly} answered
questions, and another for the estimated knowledge at the coordinates of
\textit{incorrectly} answered questions (Fig.~\ref{fig:predictions}). We then
used independent samples $t$-tests to compare the means of these distributions
of estimated knowledge.

@@

Knowledge estimates need not be limited to the content of the lectures. As
illustrated in Figure~\ref{fig:knowledge-maps}, our general approach to
estimating knowledge from a small number of quiz questions may be extended to
\textit{any} content, given its text embedding coordinate. To visualize how
knowledge ``spreads'' through text embedding space to content beyond the
lectures participants watched, we first fit a new topic model to the lectures'
sliding windows with (up to)~$k = 100$~topics.
Conceptually, increasing the number of topics used by the model functions to
increase the ``resolution'' of the embedding space, providing a greater ability to
estimate knowledge for content that is highly similar to (but not precisely the same as)
that contained in the two lectures.
Aside from increasing the number of topics from 15 to 100, all other
procedures and model parameters were carried over from the preceding analyses.
As in our other analyses, we resampled each lecture's topic trajectory to 1~Hz
and projected each question into a shared text embedding space.

@@

    \caption{\textbf{Mapping out the geometry of knowledge and learning.}
    \textbf{A. Average ``knowledge maps'' estimated using each quiz.} Each map
    displays a 2D projection of the estimated knowledge about the content
    reflected by \textit{all} regions of topic space (see
    \nameref{subsec:knowledge-maps}). The topic trajectories of the two
    lectures are indicated by dotted lines (blue: Lecture 1; green: Lecture 2),
    and the coordinates of each question are indicated by dots (light blue:
    Lecture 1-related; light green: Lecture 2-related; purple: general physics
    knowledge). Each map reflects an average across all participants. For
    individual participants' maps, see Supplementary
    Figures~\individualKnowledgeMapsA,~\individualKnowledgeMapsB,
    and~\individualKnowledgeMapsC. \textbf{B. Average ``learning maps''
    estimated between each successive pair of quizzes.} The learning maps follow
    the same general format as the knowledge maps in Panel A, but here the
    shading at each coordinate indicates the \textit{difference} between the
    corresponding coordinates in the indicated \textit{pair} of knowledge
    maps---i.e., how much the estimated knowledge ``changed'' between the two
    quizzes. Each map reflects an average across all participants. For
    individual participants' maps, see Supplementary
    Figures~\individualLearningMapsA~and~\individualLearningMapsB. \textbf{C.
    Word clouds for sampled points in topic space.} Each word cloud displays the
    weighted blend of words underlying the topic proportions represented at the
    corresponding colored star's location on the maps. In each word cloud, the
    words' relative sizes correspond to their relative weights at the starred
    location, and their colors indicate their relative weights in the
    \textit{Four Fundamental Forces} (blue) versus \textit{Birth of Stars}
    (green) lectures, on average, across all timepoints' topic vectors.}

@@

We projected the resulting 100-dimensional topic vectors (for each second of
video and each quiz question) onto a shared 2-dimensional plane (see
\nameref{subsec:knowledge-maps}). Next, we sampled points from a $100 \times
100$ grid of coordinates that evenly tiled a rectangle enclosing the 2D
projections of the videos and questions. We used
Equation~\ref{eqn:rbf-knowledge} to estimate participants' knowledge at each of
these 10,000 sampled locations, and averaged these estimates across
participants to obtain an estimated average \textit{knowledge map}
(Fig.~\ref{fig:knowledge-maps}A). Intuitively, the knowledge map constructed
from a given quiz's responses provides a visualization of how ``much''
participants knew about any content expressible by the fitted text embedding
model at the point in time when they completed that quiz.

Several features of the resulting knowledge maps are worth noting. The average
knowledge map estimated from Quiz~1 responses (Fig.~\ref{fig:knowledge-maps}A,
leftmost map) shows that participants tended to have relatively little
knowledge about any parts of the text embedding space (i.e., the shading is
relatively dark everywhere). The knowledge map estimated from Quiz~2 responses
shows a marked increase in knowledge on the left side of the map (around
roughly the same range of coordinates traversed by the \textit{Four Fundamental
Forces} lecture, indicated by the dotted blue line). In other words,
participants' estimated increase in knowledge is localized to conceptual
content that is nearby (i.e., related to) the content from the lecture they
watched prior to taking Quiz~2. This localization is non-trivial: these knowledge
estimates are informed only by the embedded coordinates of the \textit{quiz
questions}, not by the embeddings of either lecture
(see Eqn.~\ref{eqn:rbf-knowledge}). Finally, the knowledge map estimated from Quiz
3 responses shows a second increase in knowledge, localized to the region
surrounding the embedding of the \textit{Birth of Stars} lecture participants
watched immediately prior to taking Quiz~3.

Another way of visualizing these content-specific increases in knowledge
after participants viewed each lecture is displayed in
Figure~\ref{fig:knowledge-maps}B. Taking the point-by-point difference between
the knowledge maps estimated from responses to a successive pair of quizzes
yields a \textit{learning map} that describes the \textit{change} in knowledge
estimates from one quiz to the next. These learning maps highlight that the
estimated knowledge increases we observed across maps were specific to the
regions around the embeddings of each lecture, in turn.

Because the 2D projection we used to construct the knowledge and learning maps
is invertible, we may gain additional insights into these maps' meaning
by reconstructing the original high-dimensional topic vector for any location
on the map we are interested in. For example, this could serve as a useful
tool for an instructor looking to better understand which content areas a
student (or a group of students) knows well (or poorly). As a demonstration, we
show the top-weighted words from the blends of topics reconstructed from three
example locations on the maps (Fig.~\ref{fig:knowledge-maps}C): one point near
the \textit{Four Fundamental Forces} embedding (yellow), a second point near
the \textit{Birth of Stars} embedding (orange), and a third point
between the two lectures' embeddings (pink). As shown in the word clouds in the
panel, the top-weighted words at the example coordinate near the \textit{Four
Fundamental Forces} embedding tended to be weighted more heavily by the topics
expressed in that lecture. Similarly, the top-weighted words at the example
coordinate near the \textit{Birth of Stars} embedding tended to be weighted
more heavily by the topics expressed in \textit{that} lecture. And the
top-weighted words at the example coordinate between the two lectures'
embeddings show a roughly even mix of words most strongly associated with each
lecture.
